---
title: "STAT 450/460"
author: 'Handout 2: Continuous Random Variables'
date: "Fall 2016"
output:  pdf_document

---

# Chapter 4: Continuous random variables 

To define a continuous random variable, we must first define the *cumulative distribution function*, often notated as $F(y)$.  A CDF can be defined for **any** random variable $Y$.

> ***Definition***: A function, $F(y) = P(Y\leq y)$, $y\in \mathcal{R}$, is a CDF if and only if:

> 1.  $\lim_{y\rightarrow -\infty} F(y) = 0$, and $\lim_{y\rightarrow \infty}F(y) = 1$
> 2.  $F(y)$ is nondecreasing: $F(y_1) \leq F(y_2)$ if $y_1 \leq y_2$
> 3. $F(y)$ is right-continuous: $\lim_{y\rightarrow y_0^+} F(y) = F(y_0)$

Recall from handout 1; $Y\equiv$ number of heads out of 3 flips.  The pmf was:


 y  | p(y) 
--- | ---
0   | 1/8 = 0.125
1   | 3/8 = 0.375
2   | 3/8 = 0.375
3   | 1/8 = 0.125

The CDF would look like the following:



\newpage

> ***Definition***: A random variable $Y$ with distribution function $F(y)$ is said to be *continuous* if $F(y)$ is continuous, for $-\infty < y < \infty$.  


A typical CDF for a continuous random variable:  

\vspace{3in}

The derivative of $F(y)$ (**if it exists**) is also extremely important for theoretical statistics.  The derivative (if it exists) is notated by $f(y)$ and is called the **probability density function** (pdf) of $Y$.

> ***Definition***  Let $F(y)$ be the distribution function for a continuous random variable $Y$. Then $f(y)$, given by
>
$$f(y) = \frac{dF(y)}{dy} = F'(y)$$ 
>
> is called the **probability density function** (pdf) for the random variable $Y$ wherever $F'(y)$ exists.  

It follows from this definition, and from the Fundamental Theorem of Calculus, that:

$$F(y) = \int_{-\infty}^y f(t) dt = P(Y\leq y).$$

***Properties of a pdf***: If $f(y)$ is a probability density function for a continuous random variable, then:

1. $f(y) \geq 0$ for all $y$; $-\infty < y < \infty$ 
2. $\int_{-\infty}^{\infty} f(y) dy = 1$ 
3. $P(Y=y) = 0$: $\int_y^y f(t) dt = 0$
4. $P(a\leq Y \leq b) = \int_a^b f(y) dy$; note that the inclusion of endpoints ($\leq$ vs $<$) doesn't matter for continuous random variables.  



\newpage

**Expectation:** 

  * $E(Y) = \int_{-\infty}^{\infty} yf(y) dy$ 
  * $E(g(Y)) = \int_{-\infty}^{\infty} g(y)f(y) dy$
  * $Var(Y) = \int_{-\infty}^{\infty} (y-\mu)^2f(y) dy = E(Y^2)-E(Y)^2$  

**MGFs:**

  * $M_Y(t) = E(e^{tY}) = \int_{-\infty}^{\infty} e^{ty}f(y) dy$ for $t\in \{-h,h\}$


###Common Continuous Random Variables

* Uniform
* Exponential
* Gamma (survival times)
* Weibull  (survival analysis)
* Rayleigh (physics)
* Maxwell (physics)
* Normal (!!!)
* Cauchy - the straw man of pdfs
* Beta - used to model probabilities; $y\in [0,1]$

**Example**

$$f(y) = \left\{\begin{array}
{ll}
ky^2(2-y) & 0\leq y \leq 2 \\
0 & otherwise\\
\end{array}\right.
$$

For this pdf, the *support* is $[0,2]$: the support is defined to be the region where $f(y) >0$.  

**Tasks** 

a) Find $k$ such that $f(y)$ is a pdf, and graph the pdf.
b) Find the CDF, $F(y)$, and graph it.
c) Find $p(1<Y<2)$.
d) Find $E(Y)$. 
e) Find $Var(Y)$. 
f) Find the median, $m$. 

\newpage
a)  Find $k$ such that $f(y)$ is a pdf, and graph the pdf.

\vspace{1.5in}

R code to plot pdf:
```{r,eval=FALSE}
f.y <- function(y) {
  pdf <-  ifelse( y < 0 | y > 2,0, 0.75*y^2*(2-y))
  return(pdf)
}
yvals <- seq(-1,3,length=300)
mydata <- data.frame(y =  yvals, height= f.y(yvals))
library(ggplot2)
ggplot(aes(x=y, y = height), data = mydata) + geom_line(size=2) +
    ylab('f(y)') + ggtitle('probability density function')
```

b) Find the CDF, $F(y)$, and graph it.
\vspace{1.8in}


R code to plot CDF:

```{r,eval=FALSE}
#Have to modify since we have three regions to define instead of just 2:
F.y <- function(y) {
  CDF <- rep(NA,length(y))
  region1 <- which(y < 0)
  region2 <- which(0<=y & y <=2)
  region3 <- which(y>2)
  CDF[region1] <- 0
  CDF[region2] <- 0.5*y[region2]^3-3*y[region2]^4/16
  CDF[region3] <- 1
  return(CDF)
}
yvals <- seq(-1,3,length=300)
mydata <- data.frame(y =  yvals, height= F.y(yvals))
library(ggplot2)
ggplot(aes(x=y, y = height), data = mydata) + geom_line(size=2) +
    ylab('F(y)') + ggtitle('Cumulative Distribution Function')
```

\newpage
c) Find $p(1<Y<2)$.


\vspace{2.5in}


d) Find $E(Y)$.

\vspace{2.5in}

e) Find $Var(Y)$.  

\newpage

f) Find $m$, the median of $Y$.   


\vspace{2.5in}

Solving this using R:

```{r,fig.width=4,fig.height=3,warning=FALSE}
library(ggplot2)
integral <- function(m) {
  tosolve <- .5*m^3-3*m^4/16-0.5
  return(tosolve)
}
mvals <- seq(0,2,l=100)
newdat <- data.frame( m = seq(0,2,l=100), value = integral(mvals))
ggplot(aes(x=m,y=value),data=newdat) + geom_line()
#Kind of looks like the median is around 1.25.  
#Let's find the exact root using the R function uniroot():
uniroot(integral,interval=c(0,2))
```






